---
title: "TCGA Analysis Workshop - Automated Workflow"
subtitle: "Exploring Gene Expression in Cancer"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    self-contained: true
  pdf:
    toc: true
    documentclass: article
    fontsize: 11pt
    geometry: margin=1in
editor: visual
author: "Workshop Participant"
date: "`r Sys.time()`"
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

# Workshop Overview

This workshop will guide you through analyzing cancer genomics data from TCGA (The Cancer Genome Atlas) using R. By the end, you'll be able to:

-   Download gene expression, clinical and mutation data for any TCGA cancer type
-   Analyze expression of your genes of interest
-   Visualize differences between tumor and normal samples
-   Perform differential expression analysis, mutation oncoplots and correlation plots
-   Create publication-quality plots

**Key Features:** Most of the workflow runs automatically! You only need to specify your genes of interest and cancer type at the beginning.

------------------------------------------------------------------------

# Workshop Learning Objectives - Checklist

By completing this workshop, you should be able to:

-   [ ] Download TCGA data using TCGAbiolinks
-   [ ] Process and normalize RNA-seq count data
-   [ ] Perform differential expression analysis
-   [ ] Create publication-quality visualizations
-   [ ] Interpret survival analysis results
-   [ ] Understand gene co-expression patterns
-   [ ] Analyze mutation data and create oncoplots
-   [ ] Explore clinical variables and their impact
-   [ ] Modify the workflow for your own research questions

------------------------------------------------------------------------

# PART 1: Setup & Configuration

## 1.1 Install Required Packages (ONE-TIME ONLY)

**Run this section only ONCE when setting up R for the first time**

```{r install-packages, eval=FALSE}
# Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install Bioconductor packages
BiocManager::install(c(
  "TCGAbiolinks",
  "SummarizedExperiment", 
  "edgeR",
  "DESeq2",
  "org.Hs.eg.db",
  "maftools"
), ask = FALSE, update = FALSE)

# Install CRAN packages
install.packages(c(
  "dplyr", "ggplot2", "pheatmap", "RColorBrewer",
  "survival", "survminer", "ggpubr", "tibble", 
  "Hmisc", "patchwork", "gridExtra", "reactable"
))

```

\

## 1.2 Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Load all required libraries
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)
library(ggplot2)
library(edgeR)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(survival)
library(survminer)
library(ggpubr)
library(tibble)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(maftools)
library(patchwork)
library(gridExtra)
library(reactable)

# Set working directory for saving results or create a new project
# MODIFY THIS to your preferred location
# MG - remove setwd and talk through set new project in workshop 
# setwd("~/TCGA_Workshop_Results")
# 
# # Create directory if it doesn't exist
# dir.create("~/TCGA_Workshop_Results", showWarnings = FALSE)

cat("All libraries loaded successfully!\n")
cat("Results will be saved to:", getwd(), "\n")
```

------------------------------------------------------------------------

# PART 2: Customize Your Analysis

## 2.1 Define Your Research Question

**EDIT THESE TWO PARAMETERS TO CUSTOMIZE YOUR ANALYSIS:**

```{r parameters}
# ==============================================================================
# ⭐ MODIFY THESE PARAMETERS FOR YOUR ANALYSIS ⭐
# ==============================================================================

# 1. Specify your genes of interest (Gene Symbols)
genes_of_interest <- c(
  "TP53",      # Tumor suppressor
  "KRAS",      # Oncogene
  "EGFR",      # Growth factor receptor
  "MYC",       # Transcription factor
  "BRCA1",     # DNA repair
  "PTEN"       # Tumor suppressor
)

# Note: When editing genes, be mindful of the whitespace and capitalisation. If you are unsure, check your gene symbol at https://www.ncbi.nlm.nih.gov/datasets/gene/

# 2. Specify your TCGA cancer type
# Common options:
# "TCGA-BRCA" = Breast Cancer
# "TCGA-LUAD" = Lung Adenocarcinoma
# "TCGA-STAD" = Stomach Adenocarcinoma
# "TCGA-COAD" = Colon Adenocarcinoma
# "TCGA-PRAD" = Prostate Adenocarcinoma
# "TCGA-THCA" = Thyroid Carcinoma
# "TCGA-LIHC" = Liver Hepatocellular Carcinoma
# "TCGA-KIRC" = Kidney Renal Clear Cell Carcinoma

cancer_type <- "TCGA-PAAD"

# ==============================================================================
# ⚠️ DO NOT MODIFY BELOW THIS LINE - AUTOMATION STARTS HERE ⚠️
# ==============================================================================

cat("\n=== ANALYSIS CONFIGURATION ===\n")
cat("Genes of Interest:", paste(genes_of_interest, collapse = ", "), "\n")
cat("Cancer Type:", cancer_type, "\n")
```

------------------------------------------------------------------------

# PART 3: Automated Data Download & Processing

## 3.1 Query and Download Gene Expression Data

```{r query-download, include=FALSE}
cat("Step 1: Querying TCGA database for", cancer_type, "...\n")

# Query for RNA-seq gene expression data
query_exp <- GDCquery(
  project = cancer_type,
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts",
  sample.type = c("Primary Tumor", "Solid Tissue Normal")
)

cat("Query completed!\n")
cat("   Samples found:", length(query_exp$results[[1]]$id), "\n\n")

cat("Step 2: Downloading data (this may take several minutes)...\n")
GDCdownload(query_exp, method = "api")
```

```{r}

cat("Step 3: Preparing data for analysis...\n")
data_exp <- GDCprepare(query_exp)

cat("Total genes:", nrow(data_exp), "\n")
cat("Total samples:", ncol(data_exp), "\n\n")
```

## 3.2 Extract and Process Count Matrix

```{r process-counts}
cat("Processing count matrix...\n")

# Extract raw counts
counts_matrix <- assay(data_exp)

# Clean ENSEMBL IDs (remove version numbers)
rownames(counts_matrix) <- gsub("\\..*", "", rownames(counts_matrix))

cat("Count matrix extracted!\n")
cat("Dimensions:", nrow(counts_matrix), "genes ×", ncol(counts_matrix), "samples\n\n")
```

## 3.3 Map Gene Symbols to ENSEMBL IDs

```{r map-genes}
cat("Mapping gene symbols to ENSEMBL IDs...\n")

# Map gene symbols to ENSEMBL IDs
mapping <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = genes_of_interest,
  columns = c("ENSEMBL", "SYMBOL"),
  keytype = "SYMBOL"
)

# Remove duplicates and missing values
mapping <- mapping[!is.na(mapping$ENSEMBL), ]
mapping <- mapping[!duplicated(mapping$SYMBOL), ]

# Check which genes are available in the dataset
ensembl_ids <- rownames(counts_matrix)
available <- mapping[mapping$ENSEMBL %in% ensembl_ids, ]
missing <- setdiff(genes_of_interest, available$SYMBOL)

cat("\n Gene Mapping Results:\n")
cat("   Genes found:", length(available$SYMBOL), "/", length(genes_of_interest), "\n")
if (length(missing) > 0) {
  cat("Genes not found:", paste(missing, collapse = ", "), "\n")
} else {
  cat("All genes found in dataset!\n")
}

# Update genes of interest to only include available genes
genes_of_interest <- available$SYMBOL
cat("\n Proceeding with", length(genes_of_interest), "genes\n\n")
```

## 3.4 Perform Differential Expression Analysis (ALL GENES)

```{r differential-expression}
cat("Running Differential Expression Analysis...\n")
cat("This analyzes ALL genes in the dataset to find tumor vs normal differences\n\n")

# Extract metadata
metadata <- colData(data_exp) %>% as.data.frame()

# Create sample type labels
metadata$sample_type <- ifelse(
  metadata$shortLetterCode == "TP", "Tumor",
  ifelse(metadata$shortLetterCode == "NT", "Normal", NA)
)

# Remove samples without proper labels
metadata <- metadata[!is.na(metadata$sample_type), ]
counts_for_de <- counts_matrix[, metadata$barcode]

# Ensure sample alignment
metadata <- metadata[match(colnames(counts_for_de), metadata$barcode), ]

cat("Samples for analysis:\n")
cat("Tumor samples:", sum(metadata$sample_type == "Tumor"), "\n")
cat("Normal samples:", sum(metadata$sample_type == "Normal"), "\n\n")

# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = counts_for_de,
  colData = metadata,
  design = ~ sample_type
)

# Run DESeq2
cat("Running DESeq2 normalization and testing...\n")
dds <- DESeq(dds)

# Get results
de_results <- results(dds, contrast = c("sample_type", "Tumor", "Normal"))
de_results <- as.data.frame(de_results)

# Add gene symbols
de_results$ensembl <- rownames(de_results)
gene_mapping_all <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = de_results$ensembl,
  columns = c("SYMBOL", "ENSEMBL"),
  keytype = "ENSEMBL"
)
gene_mapping_all <- gene_mapping_all[!duplicated(gene_mapping_all$ENSEMBL), ]
de_results <- merge(de_results, gene_mapping_all, by.x = "ensembl", by.y = "ENSEMBL", all.x = TRUE)

# Sort by adjusted p-value
de_results <- de_results[order(de_results$padj, na.last = TRUE), ]

cat("Differential Expression Analysis Complete!\n")
cat("Total genes tested:", nrow(de_results), "\n")
cat("Significant genes (padj < 0.05):", sum(de_results$padj < 0.05, na.rm = TRUE), "\n\n")

# Save results
write.csv(de_results, "DE_results_all_genes.csv", row.names = FALSE)
cat("Full results saved to: DE_results_all_genes.csv\n\n")

# Display top 20 differentially expressed genes
cat("Top 20 Differentially Expressed Genes:\n")
top_genes <- head(de_results[!is.na(de_results$SYMBOL), 
                             c("SYMBOL", "log2FoldChange", "padj")], 20)
rownames(top_genes) <- NULL
print(top_genes)
```

## 3.5 Normalize Expression Data (logCPM) and extract genes of interest for plotting

```{r normalize}
cat("\n Normalizing expression data for visualization...\n")
# Normalize using edgeR
dge <- DGEList(counts = counts_matrix)
dge <- calcNormFactors(dge)
logCPM <- cpm(dge, log = TRUE, prior.count = 1)

cat("Normalization complete!\n")
cat("Expression matrix:", nrow(logCPM), "genes ×", ncol(logCPM), "samples\n\n")

# Subset to genes of interest
logCPM_subset <- logCPM[available$ENSEMBL, ]
rownames(logCPM_subset) <- available$SYMBOL

cat("Subset Expression matrix:", nrow(logCPM_subset), "genes ×", ncol(logCPM_subset), "samples\n\n")


```

## 3.6 Extract DE Results for Your Genes of Interest

```{r extract-interest-genes}
cat("\n Extracting results for your genes of interest...\n")

# Filter for genes of interest
genes_de_results <- de_results[de_results$SYMBOL %in% genes_of_interest, 
                               c("SYMBOL", "log2FoldChange", "pvalue", "padj")]
genes_de_results <- genes_de_results[order(genes_de_results$padj), ]
rownames(genes_de_results) <- NULL

cat("\n Differential Expression Results for Your Genes:\n")
print(genes_de_results)

# Save results for genes of interest
write.csv(genes_de_results, "DE_results_genes_of_interest.csv", row.names = FALSE)
cat("\n Results saved to: DE_results_genes_of_interest.csv\n")
```


## 3.7 Prepare Metadata

```{r prepare-metadata}
cat("Preparing sample metadata...\n")

# Extract and clean metadata
meta <- colData(data_exp) %>% as.data.frame()

# Standardize sample identifiers
colnames(logCPM_subset) <- substr(colnames(logCPM_subset), 1, 16)
meta$barcode <- substr(meta$barcode, 1, 16)

# Keep only common samples
common_samples <- intersect(colnames(logCPM_subset), meta$barcode)
logCPM_subset <- logCPM_subset[, common_samples]
meta <- meta[meta$barcode %in% common_samples, ]

# Align metadata to expression matrix
meta <- meta[match(colnames(logCPM_subset), meta$barcode), ]

# Verify alignment
if (all(colnames(logCPM_subset) == meta$barcode)) {
  cat("Metadata aligned with expression data!\n\n")
} else {
  cat("Warning: Metadata alignment issue!\n\n")
}
```

## 3.8 Prepare Clinical Data for Survival Analysis

```{r clinical-data}
cat("Preparing clinical data for survival analysis...\n")

# Create survival dataframe
clinical_clean <- meta %>%
  dplyr::select(patient = submitter_id, vital_status, 
                days_to_death, days_to_last_follow_up, shortLetterCode) %>%
  dplyr::mutate(
    time = ifelse(!is.na(days_to_death), days_to_death, days_to_last_follow_up),
    status = ifelse(vital_status == "Dead", 1, 0)
  ) %>%
  dplyr::distinct(patient, .keep_all = TRUE) %>%
  filter(!is.na(time), time > 0)

cat("Clinical data prepared!\n")
cat("Patients with survival data:", nrow(clinical_clean), "\n")
cat("Events (deaths):", sum(clinical_clean$status == 1), "\n\n")
```

## 3.9 Save Analysis Workspace

```{r save-workspace}
cat("Saving analysis workspace...\n")

# Save all objects for later use
save.image(file = paste0(cancer_type,"_analysis_workspace.RData"))

cat("Workspace saved to:", paste0(cancer_type,"_analysis_workspace.RData"), "\n")
cat("You can reload this anytime with: load('", cancer_type,"_analysis_workspace.RData')\n\n")
```

------------------------------------------------------------------------

# PART 4: Visualization & Results

## 4.1 Volcano Plot - Overview of All DE Genes

```{r volcano-plot, fig.width=10, fig.height=8}
cat("Creating Volcano Plot...\n")

# Prepare data for volcano plot
volcano_data <- de_results %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    significance = case_when(
      padj < 0.05 & abs(log2FoldChange) > 1 ~ "Significant",
      TRUE ~ "Not Significant"
    )
  )

# Create volcano plot
p_volcano <- ggplot(volcano_data, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.5, size = 1) +
  geom_point(data = volcano_data[!is.na(volcano_data$gene_label), ],
             aes(x = log2FoldChange, y = -log10(padj)),
             color = "red", size = 3, shape = 21, fill = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "gray60")) +
  labs(
    title = paste("Volcano Plot:", cancer_type, "Tumor vs Normal"),
    subtitle = "Your genes of interest are highlighted in red",
    x = "Log2 Fold Change (Tumor vs Normal)",
    y = "-Log10 Adjusted P-value",
    color = "Significance"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40")
  )

print(p_volcano)

ggsave("volcano_plot.pdf", p_volcano, width = 10, height = 8)
cat("Volcano plot saved to: volcano_plot.pdf\n\n")
```

## 4.2 Boxplots - Expression in Tumor vs Normal

```{r boxplots, fig.width=12, fig.height=10}
cat("Creating expression boxplots for your genes...\n")

# Prepare data for boxplots
expr_long <- logCPM_subset %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  left_join(
    meta %>% dplyr::select(sample = barcode, group = shortLetterCode),
    by = "sample"
  ) %>%
  filter(group %in% c("TP", "NT")) %>%
  mutate(group = factor(group, levels = c("NT", "TP"),
                       labels = c("Normal", "Tumor")))

# Create boxplots
boxplot_list <- list()

for (gene in genes_of_interest) {
  gene_data <- expr_long %>% filter(gene == !!gene)
  
  # Statistical test
  stat_test <- wilcox.test(
    expression ~ group,
    data = gene_data
  )
  
  # Get DE results for this gene
  gene_de <- genes_de_results[genes_de_results$SYMBOL == gene, ]
  fc_text <- ifelse(nrow(gene_de) > 0, 
                   sprintf("FC=%.2f, padj=%.2e", 
                          2^gene_de$log2FoldChange, 
                          gene_de$padj),
                   "")
  
  p <- ggplot(gene_data, aes(x = group, y = expression, fill = group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 1) +
    scale_fill_manual(values = c("Normal" = "#4DBBD5", "Tumor" = "#E64B35")) +
    labs(
      title = gene,
      subtitle = fc_text,
      x = "",
      y = "Expression (log2 CPM)"
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 9, color = "gray40")
    ) +
    stat_compare_means(
      method = "wilcox.test",
      label = "p.format",
      size = 3.5
    )
  
  boxplot_list[[gene]] <- p
}

# Arrange plots
combined_boxplots <- wrap_plots(boxplot_list, ncol = 2)
print(combined_boxplots)

ggsave("expression_boxplots.pdf", combined_boxplots, width = 12, height = 10)
cat("Coxplots saved to: expression_boxplots.pdf\n\n")
```

## 4.3 Heatmap - Expression Pattern Across Samples

```{r heatmap, fig.width=12, fig.height=8}
cat("Creating expression heatmap...\n")

# Prepare expression matrix
expr_mat <- logCPM_subset

# Prepare sample annotation
sample_info <- meta %>%
  filter(barcode %in% colnames(expr_mat)) %>%
  dplyr::select(sample = barcode, shortLetterCode) %>%
  mutate(shortLetterCode = factor(shortLetterCode,
                                 levels = c("NT", "TP"),
                                 labels = c("Normal", "Tumor"))) %>%
  arrange(shortLetterCode)

# Reorder samples
expr_mat <- expr_mat[, sample_info$sample]

# Create annotation
ann_col <- data.frame(
  Group = sample_info$shortLetterCode,
  row.names = sample_info$sample
)

# Create heatmap
pheatmap(
  expr_mat,
  annotation_col = ann_col,
  show_rownames = TRUE,
  show_colnames = FALSE,
  color = colorRampPalette(rev(brewer.pal(n = 11, name = "RdYlBu")))(100),
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  clustering_method = "ward.D2",
  scale = "row",
  annotation_colors = list(Group = c("Normal" = "#4DBBD5", "Tumor" = "#E64B35")),
  main = paste("Gene Expression Heatmap:", cancer_type),
  fontsize = 10,
  fontsize_row = 8
)

cat("Heatmap created!\n\n")
```

## 4.4 Survival Analysis - Tumor Samples Only

```{r survival, fig.width=12, fig.height=10}
cat("Creating survival curves...\n")

# Prepare patient IDs
patient_ids <- substr(colnames(logCPM_subset), 1, 12)

# List to store plots
survival_plots <- list()

for (gene in genes_of_interest) {
  
  # Expression data
  exp_df <- data.frame(
    patient = patient_ids,
    expr = as.numeric(logCPM_subset[gene, ])
  ) %>%
    group_by(patient) %>%
    summarise(expr = mean(expr, na.rm = TRUE))
  
  # Merge with clinical data
  merged <- inner_join(clinical_clean, exp_df, by = "patient")
  
  # High vs Low expression (median split)
  merged$expr_group <- ifelse(
    merged$expr >= median(merged$expr, na.rm = TRUE),
    "High", "Low"
  )
  
  # Keep only tumor samples
  merged <- merged[merged$shortLetterCode == "TP", ]
  
  # Skip if insufficient data
  if (length(unique(merged$expr_group)) < 2) {
    cat("Skipping", gene, "- insufficient variation\n")
    next
  }
  
  # Survival analysis
  fit <- survfit(Surv(time, status) ~ expr_group, data = merged)
  
  # Cox proportional hazards
  cox <- coxph(Surv(time, status) ~ expr_group, data = merged)
  cox_summary <- summary(cox)
  
  # Extract HR
  hr <- round(cox_summary$coefficients[, "exp(coef)"], 2)
  ci_lower <- round(cox_summary$conf.int[, "lower .95"], 2)
  ci_upper <- round(cox_summary$conf.int[, "upper .95"], 2)
  hr_text <- paste0("HR = ", hr, " (95% CI: ", ci_lower, "–", ci_upper, ")")
  
  # Create survival plot
  p_surv <- ggsurvplot(
    fit,
    data = merged,
    pval = TRUE,
    title = paste0(gene, " Expression & Survival (Tumor Patients)"),
    risk.table = TRUE,
    risk.table.height = 0.25,
    palette = c("#E64B35", "#4DBBD5"),
    legend.labs = c("High", "Low"),
    ggtheme = theme_bw(),
    legend.title = "Expression",
    xlab = "Time (days)"
  )
  
  # Add HR annotation
  p_surv$plot <- p_surv$plot +
    annotate("text",
            x = max(merged$time, na.rm = TRUE) * 0.6,
            y = 0.1,
            label = hr_text,
            size = 3.5)
  
  survival_plots[[gene]] <- p_surv
}

# Display plots
if (length(survival_plots) > 0) {
  arrange_ggsurvplots(survival_plots, ncol = 2)
  cat("Survival analysis complete!\n\n")
} else {
  cat("No survival plots generated\n\n")
}
```

## 4.5 Correlation Analysis - Gene Co-expression

```{r correlation, fig.width=10, fig.height=8}
cat("Analyzing gene correlations...\n")

# Prepare expression matrix (tumor samples only)
expr_mat <- t(logCPM_subset[genes_of_interest, ])
tumor_samples <- meta$barcode[meta$shortLetterCode == "TP"]
tumor_samples <- tumor_samples[tumor_samples %in% rownames(expr_mat)]
expr_tumor <- expr_mat[tumor_samples, , drop = FALSE]

# Calculate Spearman correlation
library(Hmisc)
res <- rcorr(as.matrix(expr_tumor), type = "spearman")
cor_mat <- res$r
p_mat <- res$P

# Create significance asterisks
sig_stars <- matrix("", nrow = nrow(p_mat), ncol = ncol(p_mat))
sig_stars[p_mat <= 0.001] <- "***"
sig_stars[p_mat <= 0.01 & p_mat > 0.001] <- "**"
sig_stars[p_mat <= 0.05 & p_mat > 0.01] <- "*"

# Combine correlations and significance
display_mat <- matrix("", nrow = nrow(cor_mat), ncol = ncol(cor_mat))
for (i in 1:nrow(cor_mat)) {
  for (j in 1:ncol(cor_mat)) {
    display_mat[i, j] <- paste0(sprintf("%.2f", cor_mat[i, j]), sig_stars[i, j])
  }
}

# Create heatmap
pheatmap(
  cor_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  display_numbers = display_mat,
  number_color = "black",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = paste("Gene Correlation Heatmap (Tumor Samples):", cancer_type),
  fontsize = 10
)

cat("Correlation analysis complete!\n\n")
cat("Note: * p≤0.05, ** p≤0.01, *** p≤0.001\n\n")
```

## 4.6 Mutation Analysis - Download and Visualize

```{r mutation-download, include=FALSE}
cat("Downloading mutation data...\n")

# Query mutation data
query_mut <- GDCquery(
  project = cancer_type,
  data.category = "Simple Nucleotide Variation",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

# Download mutation data
GDCdownload(query_mut, method = "api")

# Prepare mutation data
stad_maf <- GDCprepare(query_mut)
maf_object <- read.maf(stad_maf)

cat("Mutation data downloaded!\n\n")
```

### Mutation Landscape - Oncoplot

```{r mutation-oncoplot, fig.width=14, fig.height=10}
cat("Creating mutation oncoplot...\n")

# Plot mutations for genes of interest
cat("\n Mutations in Your Genes of Interest:\n")
oncoplot(
  maf = maf_object,
  genes = genes_of_interest
)


cat("\n Oncoplot shows mutation types and frequencies across samples\n")
cat("   - Rows: Your genes of interest\n")
cat("   - Columns: Patient samples\n")
cat("   - Colors: Different mutation types (missense, nonsense, frameshift, etc.)\n\n")
```

### Top Mutated Genes in Dataset

```{r mutation-top-genes, fig.width=14, fig.height=10}
cat("Top 20 Most Mutated Genes in", cancer_type, ":\n")

oncoplot(
  maf = maf_object,
  top = 20
)


cat("\n This shows the most frequently mutated genes in the entire cohort\n\n")
```

### Mutation Summary Statistics

```{r mutation-summary}
cat("Mutation Summary Statistics:\n\n")

# Get summary
maf_summary <- getSampleSummary(maf_object)

# Get gene summary
gene_summary <- getGeneSummary(maf_object)

# Filter for genes of interest
genes_mut_summary <- gene_summary[gene_summary$Hugo_Symbol %in% genes_of_interest, ]

if (nrow(genes_mut_summary) > 0) {
  cat("\n\nMutation Statistics for Your Genes:\n")
  print(genes_mut_summary[, c("Hugo_Symbol", "total", "MutatedSamples", "AlteredSamples")])
} else {
  cat("\n\n⚠️ None of your genes of interest have mutations in this dataset\n")
}
```

#### Table: Mutation types in sample sorted by highest total mutations

```{r mutation-summary-reactable}
# Display mutated samples sorted by most mutations
reactable(maf_summary, 
          defaultSortOrder = "desc", 
          defaultPageSize = 5, 
          striped = TRUE)

```

## 4.7 Clinical Variables Analysis

```{r clinical-variables}
cat("Analyzing clinical variables...\n")

# Extract detailed clinical data
clinical_data <- meta %>%
  dplyr::select(
    patient = submitter_id,
    age = age_at_index,
    gender,
    race,
    ethnicity,
    vital_status,
    tumor_stage = ajcc_pathologic_stage,
    sample_type = shortLetterCode
  ) %>%
  distinct(patient, .keep_all = TRUE)

cat("\n Clinical Data Summary:\n")
cat("   Total patients:", nrow(clinical_data), "\n\n")

# Age distribution
cat("Age Statistics:\n")
age_stats <- summary(clinical_data$age)
print(age_stats)

# Gender distribution
cat("\n\nGender Distribution:\n")
gender_table <- table(clinical_data$gender, useNA = "ifany")
print(gender_table)

# Race distribution
cat("\n\nEthnicity Distribution:\n")
race_table <- table(clinical_data$race, useNA = "ifany")
print(race_table)

# Tumor stage distribution
cat("\n\nTumor Stage Distribution:\n")
stage_table <- table(clinical_data$tumor_stage, useNA = "ifany")
print(stage_table)

# Vital status
cat("\n\nVital Status:\n")
vital_table <- table(clinical_data$vital_status, useNA = "ifany")
print(vital_table)

cat("\n")
```

### Clinical Variables Visualization

```{r clinical-viz, fig.width=12, fig.height=8}
cat("Creating clinical variable visualizations...\n")

# Prepare data for plotting (tumor samples only)
tumor_clinical <- clinical_data %>%
  filter(!is.na(age), !is.na(gender))

# Create plots
p1 <- ggplot(tumor_clinical, aes(x = age)) +
  geom_histogram(bins = 30, fill = "#16A085", color = "white") +
  labs(title = "Age Distribution", x = "Age at Diagnosis", y = "Count") +
  theme_bw()

p2 <- ggplot(tumor_clinical, aes(x = gender, fill = gender)) +
  geom_bar() +
  scale_fill_manual(values = c("female" = "#E74C3C", "male" = "#3498DB")) +
  labs(title = "Gender Distribution", x = "Gender", y = "Count") +
  theme_bw() +
  theme(legend.position = "none")

# Combine plots
combined_clinical <- p1 + p2
print(combined_clinical)

cat("Clinical visualizations created!\n\n")
```

### Survival by Clinical Variable

```{r survival-clinical, fig.width=10, fig.height=6}
cat("Survival analysis by clinical variables...\n")


# Prepare survival data with clinical variables
surv_clinical <- clinical_clean %>%
  left_join(clinical_data, by = "patient") %>%
  filter(!is.na(time), time > 0, !is.na(gender))

# Survival by gender
if (length(unique(surv_clinical$gender)) > 1) {
  fit_gender <- survfit(Surv(time, status) ~ gender, data = surv_clinical)
  
  p_gender <- ggsurvplot(
    fit_gender,
    data = surv_clinical,
    pval = TRUE,
    title = paste("Survival by Gender:", cancer_type),
    risk.table = TRUE,
    risk.table.height = 0.25,
    palette = c("#E74C3C", "#3498DB"),
    legend.labs = c("Female", "Male"),
    ggtheme = theme_bw(),
    xlab = "Time (days)"
  )
  
  print(p_gender)
  cat("Gender-based survival analysis complete!\n\n")
} else {
  cat("Insufficient gender variation for survival analysis\n\n")
}
```

# PART 5: Summary & Next Steps

## 5.1 Analysis Summary

```{r summary}
cat("RESULTS SUMMARY:\n")
cat("  Cancer Type:", cancer_type, "\n")
cat("  Genes Analyzed:", length(genes_of_interest), "\n")
cat("  Samples Analyzed:", ncol(logCPM), "\n")
cat("  - Tumor:", sum(meta$shortLetterCode == "TP"), "\n")
cat("  - Normal:", sum(meta$shortLetterCode == "NT"), "\n\n")
cat("  Rdata file:", paste0(cancer_type, "_analysis_workspace.RData\n\n"))
```

### FILES CREATED:

1. DE_results_all_genes.csv

2. DE_results_genes_of_interest.csv

3. volcano_plot.pdf

4. expression_boxplots.pdf


### ANALYSES PERFORMED:

✓ Differential Expression (ALL genes)

✓ Volcano Plot (genome-wide overview)

✓ Expression Boxplots (tumor vs normal)

✓ Expression Heatmap (clustered samples)

✓ Survival Analysis (by expression level)

✓ Gene Correlation Analysis

✓ Mutation Analysis (oncoplot & statistics)

✓ Clinical Variables Analysis

✓ Survival by Clinical Variables

------------------------------------------------------------------------

## 5.2 Interpretation Guide

**Understanding Your Results:**

1.  **Volcano Plot**: Shows all genes. Genes in the upper corners (red dots) are significantly different between tumor and normal.

2.  **Boxplots**: Compare expression levels between tumor and normal samples for each gene. P-values indicate statistical significance.

3.  **Heatmap**: Shows expression patterns across all samples. Genes that cluster together have similar expression patterns.

4.  **Survival Curves**: Show whether high vs low expression of your genes affects patient survival. Lower p-values indicate stronger associations.

5.  **Correlation Heatmap**: Shows which genes are co-expressed (positive correlation) or inversely expressed (negative correlation). Spearman correlation value >0.5 (red) indicates postive correlation, <0.5 weak positive correlation, <0 negative correlation. * indicates pvalue significance <0.05. 

6.  **Mutation Oncoplot**: Displays mutation types and frequencies for your genes across patient samples. Different colors represent different mutation types (missense, nonsense, frameshift, etc.).

7.  **Clinical Variables**: Age, gender, ethnicity, and tumor stage distributions help contextualize your findings within the patient population.

8.  **Survival by Clinical Variables**: Shows how demographic and clinical factors affect patient outcomes independently of gene expression.

## 5.3 Next Steps

**To analyze different genes or cancer types:**

1.  Go back to **PART 2.1** and modify the parameters
2.  Re-run the entire document (Ctrl/Cmd + Shift + K in RStudio)
3.  All analyses will automatically update!

**To explore further:**

-   Add more genes to `genes_of_interest`
-   Try different cancer types
-   Adjust significance thresholds
-   Export plots for presentations
-   Investigate top differentially expressed genes

------------------------------------------------------------------------

**Questions? Need Help?**

-   Check the TCGA documentation: <https://docs.gdc.cancer.gov/>
-   TCGAbiolinks vignette: <http://bioconductor.org/packages/TCGAbiolinks/>
-   Email [simran.kaur\@adelaide.edu.au](mailto:simran.kaur@adelaide.edu.au){.email} or [monica.guilhaus\@adelaide.edu.au](mailto:monica.guilhaus@adelaide.edu.au){.email}
-   Or join our fortnightly Data Science Drop in Sessions.

**Good luck with your research!**
